# Bayesian Final Exam Report

> Alex Yang - zy7ts

## Problem Statement

Our company, VusageTome is seeking to understand specific ways the user interacts with the social network platform. There are four questions that this study seeks to address:

1.  **Demographics:** what demographic features are the most predictive in terms of number of engagments and the length of engagement
2.  **Time Patterns:** hwo does time of the day and the day of week impact user engagement
3.  **Superusers:** is there any users who engage significantly more than others (i.e. superuers)? If so how do we define and identify them
4.  **Superuser demographics:** do the superusers have specific demographic features?

## Executive Summary

1.  Demographics:
    -   Age is a significant predictor for engagement where older users engaging more frequently

    -   Gender and nationality doesn't show much effect on the general user engagement
2.  Time patterns:
    -   We can see highest level of user engagement on Fridays and Saturdays specifically in the evening

    -   Wednesday and Thursdays see the lowest level of user engagement
3.  Superusers:
    -   We define the superusers as the ones that have the top 5% of engagement while adjusted for demographics and time pattern predictors above

    -   The exisistence of superusers will impact the models because they introduce large amount of variability
4.  Superusers demographics:
    -   Within the superusers population we can see that male superusers have slightly higher average engagement than female superusers

    -   We also see that superusers from first nations have the highest average engagement (over 9000 minutes) compared to Atlantian and Carthaginian superusers (both averaging 6000 minutes).

## Introduction

## Methodology

1.  Model:
    -   A Bayesian Hierarchical mode was used

        -   We considered fixed effects (systematic variations that are considered constant across individuals) as vars(gender, age, nationality, day_of_week, time_of_day)

        -   We considered random effects (variations that differ from user to user) as vars(user_id)

    -   We used Hamiltonian Monte Carlo to fit the model
2.  Response family:
    -   We used Gaussian Distribution given the vars(MinPerBlack) is a continuous variable
3.  Superusers:
    -   We define superusers are the top 5% of the users as the ones who have top 5% engagement while adjusted for fixed effects

## Analysis and Results

1.  Multilevel Hyper parameters:
    -   We see a significant variability across users (which could be seen from our intercept which is 37.38). This shows that it is important for us to choose user_id as the random effect to account for individual differences
2.  Fixed effects:
    -   Age

        -   estimate: 2.11 with a credible interval of 1.83 to 2.4

        -   We estimate that the user engagement increases by approximately

## Conclusion

## Technical Details/Implementation

### Exploratory Data Analysis

```{r}
library(tidyverse)
library(ggplot2)

# Load the dataset
data <- read.csv("DS6040F24_finaldat.csv")  # Replace with your actual file path

# Inspect the structure of the dataset
str(data)

# Convert categorical columns to factors
data <- data %>%
  mutate(
    gender = as.factor(gender),
    day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
    time_of_day = factor(time_of_day, levels = c("Morning", "Midday", "Evening")),
    nationality = as.factor(nationality)
  )

# Summary statistics
summary(data)

# 1. Distribution of MinPerBlock
ggplot(data, aes(x = MinPerBlock)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  labs(
    title = "Distribution of Minutes Per Block",
    x = "Minutes Per Block",
    y = "Count"
  ) +
  theme_minimal()

# 2. Distribution of Age
ggplot(data, aes(x = age)) +
  geom_histogram(bins = 20, fill = "green", alpha = 0.7) +
  labs(
    title = "Distribution of Age",
    x = "Age",
    y = "Count"
  ) +
  theme_minimal()

# 3. Engagement (MinPerBlock) by Gender
ggplot(data, aes(x = gender, y = MinPerBlock, fill = gender)) +
  geom_boxplot() +
  labs(
    title = "Engagement (MinPerBlock) by Gender",
    x = "Gender",
    y = "Minutes Per Block"
  ) +
  theme_minimal()

# 4. Engagement (MinPerBlock) by Time of Day
ggplot(data, aes(x = time_of_day, y = MinPerBlock, fill = time_of_day)) +
  geom_boxplot() +
  labs(
    title = "Engagement (MinPerBlock) by Time of Day",
    x = "Time of Day",
    y = "Minutes Per Block"
  ) +
  theme_minimal()

# 5. Engagement by Day of the Week
ggplot(data, aes(x = day_of_week, y = MinPerBlock, fill = day_of_week)) +
  geom_boxplot() +
  labs(
    title = "Engagement (MinPerBlock) by Day of the Week",
    x = "Day of Week",
    y = "Minutes Per Block"
  ) +
  theme_minimal()

# 6. Gender Distribution
ggplot(data, aes(x = gender, fill = gender)) +
  geom_bar() +
  labs(
    title = "Gender Distribution",
    x = "Gender",
    y = "Count"
  ) +
  theme_minimal()

# 7. Nationality Distribution
ggplot(data, aes(x = nationality, fill = nationality)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Nationality Distribution",
    x = "Nationality",
    y = "Count"
  ) +
  theme_minimal()

# 8. Correlation Heatmap for Numeric Features
numeric_data <- data %>%
  select_if(is.numeric)
correlation_matrix <- cor(numeric_data, use = "complete.obs")
ggplot(data = as.data.frame(as.table(correlation_matrix)), aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(
    title = "Correlation Heatmap",
    x = "",
    y = ""
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Model Building

**Hierarchical effects:** vars(gender, age, nationality, time_of_day, day_of_week) are fixed effects because they are not specific to the user itself; while vars(user_id) has random effects since it's differences between individual users

**Variation inference and samplers:** we'll first use fit framework we'll then examine if Hamiltonian MC is more reliable.

```{r}
# Load required libraries
library(brms)

# Define the model formula
formula <- bf(
  MinPerBlock ~ gender + age + nationality + day_of_week + time_of_day + 
    (1 | user_id)  # Random effect for user-level variability
)

# Define priors
priors <- c(
  prior(normal(0, 5), class = "b"),        # Priors for fixed effects
  prior(cauchy(0, 2), class = "sd")        # Priors for random effects
)

# Fit the model using Hamiltonian Monte Carlo
fit <- brm(
  formula = formula,
  data = data,
  family = gaussian(),  # Response family for continuous data
  prior = priors,
  chains = 2,           # Number of MCMC chains
  iter = 2000,          # Total iterations
  warmup = 1000,        # Warmup iterations
  seed = 123            # Reproducibility
)

# Print the summary of the model
summary(fit)
```

### Validations

```{r}
# Posterior predictive checks
pp_check(fit)

# Plot fixed effects
plot(marginal_effects(fit))

# Model convergence diagnostics
plot(fit)
```

### Superuser Identification

```{r}
# ---- Superuser Identification ---- #

# Extract random effects for `user_id`
random_effects <- as.data.frame(ranef(fit)$user_id)

# Rename columns for clarity
colnames(random_effects) <- c("Estimate", "Est.Error", "l-95% CI", "u-95% CI")

# Add user_id as a column
random_effects <- random_effects %>%
  mutate(user_id = rownames(ranef(fit)$user_id))

# Calculate the superuser threshold: Top 5% of random intercepts
threshold <- quantile(random_effects$Estimate, 0.95)

# Identify superusers based on the threshold
superusers <- random_effects %>%
  filter(Estimate > threshold)

# Print identified superusers
print(superusers)

# ---- Analyze Demographics of Superusers ---- #

# Merge superusers with the original dataset
superuser_data <- data %>%
  filter(user_id %in% superusers$user_id)

# Summarize demographics of superusers
superuser_demographics <- superuser_data %>%
  group_by(gender, nationality, time_of_day, day_of_week) %>%
  summarize(
    avg_age = mean(age),
    avg_minutes = mean(MinPerBlock),
    count = n()
  ) %>%
  arrange(desc(avg_minutes))

# Print the summary of superuser demographics
print(superuser_demographics)

# ---- Visualization of Superuser Demographics ---- #

# Bar plot of average minutes by gender for superusers
ggplot(superuser_demographics, aes(x = gender, y = avg_minutes, fill = gender)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Average Engagement of Superusers by Gender",
    x = "Gender",
    y = "Average Minutes"
  ) +
  theme_minimal()

# Bar plot of average minutes by nationality for superusers
ggplot(superuser_demographics, aes(x = nationality, y = avg_minutes, fill = nationality)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Average Engagement of Superusers by Nationality",
    x = "Nationality",
    y = "Average Minutes"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
