# Bayesian Final Exam Report

> Alex Yang - zy7ts

## Problem Statement

Our company, VusageTome is seeking to understand specific ways the user interacts with the social network platform. There are four questions that this study seeks to address:

1.  **User engagement:**
2.  

## Executive Summary

## Introduction

## Methodology

## Analysis and Results

## Conclusion

## Technical Details/Implementation

### Exploratory Data Analysis

```{r}
library(tidyverse)
library(ggplot2)

# Load the dataset
data <- read.csv("DS6040F24_finaldat.csv")  # Replace with your actual file path

# Inspect the structure of the dataset
str(data)

# Convert categorical columns to factors
data <- data %>%
  mutate(
    gender = as.factor(gender),
    day_of_week = factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
    time_of_day = factor(time_of_day, levels = c("Morning", "Midday", "Evening")),
    nationality = as.factor(nationality)
  )

# Summary statistics
summary(data)

# 1. Distribution of MinPerBlock
ggplot(data, aes(x = MinPerBlock)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  labs(
    title = "Distribution of Minutes Per Block",
    x = "Minutes Per Block",
    y = "Count"
  ) +
  theme_minimal()

# 2. Distribution of Age
ggplot(data, aes(x = age)) +
  geom_histogram(bins = 20, fill = "green", alpha = 0.7) +
  labs(
    title = "Distribution of Age",
    x = "Age",
    y = "Count"
  ) +
  theme_minimal()

# 3. Engagement (MinPerBlock) by Gender
ggplot(data, aes(x = gender, y = MinPerBlock, fill = gender)) +
  geom_boxplot() +
  labs(
    title = "Engagement (MinPerBlock) by Gender",
    x = "Gender",
    y = "Minutes Per Block"
  ) +
  theme_minimal()

# 4. Engagement (MinPerBlock) by Time of Day
ggplot(data, aes(x = time_of_day, y = MinPerBlock, fill = time_of_day)) +
  geom_boxplot() +
  labs(
    title = "Engagement (MinPerBlock) by Time of Day",
    x = "Time of Day",
    y = "Minutes Per Block"
  ) +
  theme_minimal()

# 5. Engagement by Day of the Week
ggplot(data, aes(x = day_of_week, y = MinPerBlock, fill = day_of_week)) +
  geom_boxplot() +
  labs(
    title = "Engagement (MinPerBlock) by Day of the Week",
    x = "Day of Week",
    y = "Minutes Per Block"
  ) +
  theme_minimal()

# 6. Gender Distribution
ggplot(data, aes(x = gender, fill = gender)) +
  geom_bar() +
  labs(
    title = "Gender Distribution",
    x = "Gender",
    y = "Count"
  ) +
  theme_minimal()

# 7. Nationality Distribution
ggplot(data, aes(x = nationality, fill = nationality)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Nationality Distribution",
    x = "Nationality",
    y = "Count"
  ) +
  theme_minimal()

# 8. Correlation Heatmap for Numeric Features
numeric_data <- data %>%
  select_if(is.numeric)
correlation_matrix <- cor(numeric_data, use = "complete.obs")
ggplot(data = as.data.frame(as.table(correlation_matrix)), aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(
    title = "Correlation Heatmap",
    x = "",
    y = ""
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Model Building

**Hierarchical effects:** vars(gender, age, nationality, time_of_day, day_of_week) are fixed effects because they are not specific to the user itself; while vars(user_id) has random effects since it's differences between individual users

**Variation inference and samplers:** we'll first use fit framework we'll then examine if Hamiltonian MC is more reliable.

```{r}
# Load required libraries
library(brms)

# Define the model formula
formula <- bf(
  MinPerBlock ~ gender + age + nationality + day_of_week + time_of_day + 
    (1 | user_id)  # Random effect for user-level variability
)

# Define priors
priors <- c(
  prior(normal(0, 5), class = "b"),        # Priors for fixed effects
  prior(cauchy(0, 2), class = "sd")        # Priors for random effects
)

# Fit the model using Hamiltonian Monte Carlo
fit <- brm(
  formula = formula,
  data = data,
  family = gaussian(),  # Response family for continuous data
  prior = priors,
  chains = 2,           # Number of MCMC chains
  iter = 2000,          # Total iterations
  warmup = 1000,        # Warmup iterations
  seed = 123            # Reproducibility
)

# Print the summary of the model
summary(fit)
```

### Validations

```{r}
# Posterior predictive checks
pp_check(fit)

# Plot fixed effects
plot(marginal_effects(fit))

# Model convergence diagnostics
plot(fit)
```
